Synthesizing Random Generators via Higher-Order Representations

### 2019-01-17 paper/main.tex
WIP

\documentclass[conference]{IEEEtran}

@@

\usepackage{todonotes}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{mathdots}
\usepackage{yhmath}
\usepackage{cancel}
\usepackage{color}
\usepackage{siunitx}
\usepackage{array}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{xspace}
\usetikzlibrary{fadings}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Macros
\newcommand{\tocite}{\textbf{CITATION NEEDED}}
\newcommand{\quickcheck}{\emph{QuickCheck}\xspace}
\newcommand{\megadeth}{\emph{MegaDeTH}\xspace}
\newcommand{\dragen}{\textbf{DRAGEN}\xspace}

@@

In the previous definition, we declare \ensuremath{\Conid{Exp}} as the sum of three possible
classes of values: \ensuremath{\Conid{Val}} represents literal values, whereas \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}
represent the addition and multiplication of two integer expressions,
respectively.
%
In Haskell, \ensuremath{\Conid{Val}}, \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}} are called \emph{data constructors} and are
used to distiguish which variant of the data type we are reffering to.
%
Each data constructor is then defined as a product of zero or more types known
as \emph{fields}.
%
In particular, \ensuremath{\Conid{Val}} contains a field of type \ensuremath{\Conid{Int}}, while \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}
contain two fields of type \ensuremath{\Conid{Exp}} representing the operands of each operation.
%
Note that \ensuremath{\Conid{Exp}} is used as a field of at least one of its data constructors
(case \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}), making it a recursively defined ADT.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this reprentation, the expression ``2 + (5 * 6)'' can be encoded simply by
using \ensuremath{\Conid{Exp}} data constructors as \ensuremath{(\Conid{Add}\;(\Conid{Val}\;\mathrm{2})\;(\Conid{Mul}\;(\Conid{Val}\;\mathrm{5})\;(\Conid{Val}\;\mathrm{6})))}.
%
Furthermore, an evaluation function from \ensuremath{\Conid{Exp}}s to integer values can be defined
very idiomatically:

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}c<{\hspost}@{}}%
\column{7E}{@{}l@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{genExp}\mathbin{::}\Conid{Gen}\;\Conid{Exp}{}\<[E]%
\\
\>[B]{}\Varid{genExp}\mathrel{=}\Varid{sized}\;\Varid{gen}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{gen}\;\mathrm{0}\mathrel{=}\Conid{Val}\mathop{\langle \$ \rangle}\Varid{arbitrary}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{gen}\;\Varid{n}\mathrel{=}\Varid{oneof}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}[\mskip1.5mu {}\<[7E]%
\>[10]{}\Conid{Val}{}\<[15]%
\>[15]{}\mathop{\langle \$ \rangle}\Varid{arbitrary}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{},{}\<[7E]%
\>[10]{}\Conid{Add}{}\<[15]%
\>[15]{}\mathop{\langle \$ \rangle}\Varid{gen}\;(n{-}1){}\<[30]%
\>[30]{}\mathop{\langle \ast \rangle}\Varid{gen}\;(n{-}1){}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{},{}\<[7E]%
\>[10]{}\Conid{Mul}{}\<[15]%
\>[15]{}\mathop{\langle \$ \rangle}\Varid{gen}\;(n{-}1){}\<[30]%
\>[30]{}\mathop{\langle \ast \rangle}\Varid{gen}\;(n{-}1)\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

This random generator is defined using \quickcheck's combinator \ensuremath{\Varid{sized}\mathbin{::}(\Conid{Int}\to \Conid{Gen}\;\Varid{a})\to \Conid{Gen}\;\Varid{a}} to paremetrize the generation process up to an external
integer number known as the \emph{generation size}.
%
This parameter is used to limit the maximum amount of recursive call that this
random generator can perform.
%
When called with a positive generation size (case \ensuremath{\Varid{gen}\;\Varid{n}}), the generator is free
to randomly pick with uniform probability among any data constructor of \ensuremath{\Conid{Exp}}.
%
In the case it picks to generate a \ensuremath{\Conid{Val}} data constructor, it also generates a
random \ensuremath{\Conid{Int}} value using the standard overloaded generator \ensuremath{\Varid{arbitrary}}
(\quickcheck provides standard random generators for most base types like \ensuremath{\Conid{Int}},
\ensuremath{\Conid{Bool}}, etc.).
%
On the other hand, when it randomly picks to generate either an \ensuremath{\Conid{Add}} or a \ensuremath{\Conid{Mul}}
data constructor, it also generates independently two random subexpressions
corresponding to the data constructor fields by calling itself recursively (\ensuremath{\Varid{gen}\;(n{-}1)}), ensuring to decrease the generation size by a unit on each recursive
invocation.

@@

\quickcheck's default implementation for random \ensuremath{\Conid{Int}}s pick a number in the
interval $[-n, n]$ with uniform distribution.
%
Hence, in principle we need to recognize what is a suitable generation size,
which should be big enough for our generator be able to generate the numbers we
pattern match against to, otherwise we will not be able to generate any value to
match against \ensuremath{\Varid{foo}} clauses, leaving fragments of code completely untested.
%
Suppose then that we pick the a generation size $50$, i.e. the minimum size that
is big enough to produce an \ensuremath{\Conid{Int}} number equal to $50$.
%
Under this consideration, the probability of generating a value matching the
first clause of \ensuremath{\Varid{foo}} (and hence triggering the first error) results as follows:

\begin{align*}
  P(match(foo\#1))
  \!&&=\; &P(Add) \\
  \!&&*\; &P(Add) * 1 * (P(Val) * (1/100)) \\
  \!&&*\; &P(Add) * (P(Val) * (1/100)) * 1
\end{align*}

If we use \megadeth to automatically derive a random generator for \ensuremath{\Conid{Exp}}, we
obtain a uniform generation probability distribution over constructors, i.e.,
$P(\ensuremath{\Conid{Val}}) = P(\ensuremath{\Conid{Add}}) = P(\ensuremath{\Conid{Mul}}) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of \ensuremath{\Varid{foo}} only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
Using this tool, we can optimize the generation probabilities in order to
benefit the generation of some data constructors over the rest.
%
Considering that the first pattern matching of \ensuremath{\Varid{foo}} involves the data
constructor \ensuremath{\Conid{Add}} as the only recursive one, we can set an target generation
proportion of \ensuremath{\Conid{Add}} data constructors of, for instance, $20:1$ with respect to
the rest of the generated data constructors.
%
By doing so, the obtained distribution of values results such that
$P(match(foo\#1)) \approx 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Html}\mathrel{=}\Conid{Html}\;\Conid{String}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{head}\mathbin{::}\Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{head}\;(\Conid{Html}\;\Varid{inner}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Conid{Html}\;(\text{\ttfamily \char34 <head>\char34}\plus \Varid{inner}\plus \text{\ttfamily \char34 </head>\char34}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{body}\mathbin{::}\Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{body}\;(\Conid{Html}\;\Varid{inner}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Conid{Html}\;(\text{\ttfamily \char34 <body>\char34}\plus \Varid{t}\plus \text{\ttfamily \char34 </body>\char34}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{br}\mathbin{::}\Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{br}\mathrel{=}\Conid{Html}\;\text{\ttfamily \char34 <br~/>\char34}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{concat}\mathbin{::}\Conid{Html}\to \Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{concat}\;(\Conid{Html}\;\Varid{x})\;(\Conid{Html}\;\Varid{y})\mathrel{=}\Conid{Html}\;(\Varid{x}\plus \Varid{y}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

We believe that, if certain patterns of values are relevant enough to appear in
the codebase being tested, a practical random testing methodology should be able
to produce values satisfying such patterns in a substantial proportion.
%
The next section introduces a reprentation model that let us encode the
structure information presented here into our automatically derived random
generators in a modular and flexible way.

% Consider the following Haskell module |M| which defines combinators for the type
% |Exp|:

% \begin{code}
% module M where

% ten :: Exp
% ten = Val 10

% square :: Exp -> Exp
% square x = Mul x x

% minus :: Exp -> Exp -> Exp
% minus x y = Add x (Mul y (Val (-1)))
% \end{code}
\section{Extracting Structure} \label{sec:hrep}

In this work, we ...

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\text{HRep}_{Val}\;{}\<[16]%
\>[16]{}\Varid{a}\mathrel{=}\text{Mk}_{Val}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{Add}\;{}\<[16]%
\>[16]{}\Varid{a}\mathrel{=}\text{Mk}_{Add}\;\Varid{a}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{Mul}\;{}\<[16]%
\>[16]{}\Varid{a}\mathrel{=}\text{Mk}_{Mul}\;\Varid{a}\;\Varid{a}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

For this purpose, we employ Haskell's type classes mechanism \tocite{}.
%
We define an evaluation type class $(\Downarrow)$ which specifies how to perform
a single transformation \ensuremath{\Varid{step}} from the higher-order representation back into a
concrete value of the target:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;(\Varid{rep}\;\Downarrow\;\Varid{target})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{step}\mathbin{::}\Varid{rep}\;\Varid{target}\to \Varid{target}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Then, we can define the following overloaded instances of the \ensuremath{\Varid{step}} operation
for the canonical representations of data constructors simply by mapping each
lifted constructor back into its corresponding one.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\text{HRep}_{Val}\;\Downarrow\;\Conid{Exp})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{step}\;(\text{Mk}_{Val}\;\Varid{n})\mathrel{=}\Conid{Val}\;\Varid{n}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\text{HRep}_{Add}\;\Downarrow\;\Conid{Exp})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{step}\;(\text{Mk}_{Add}\;\Varid{x}\;\Varid{y})\mathrel{=}\Conid{Add}\;\Varid{x}\;\Varid{y}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\text{HRep}_{Mul}\;\Downarrow\;\Conid{Exp})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{step}\;(\text{Mk}_{Mul}\;\Varid{x}\;\Varid{y})\mathrel{=}\Conid{Mul}\;\Varid{x}\;\Varid{y}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

With these individual representations, we can define a type combinator
$(\oplus)$ to compose two representations into a single one:

>>>>

\documentclass[conference, fleqn]{IEEEtran}

@@

\usepackage{todonotes}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{mathdots}
\usepackage{yhmath}
\usepackage{cancel}
\usepackage{color}
\usepackage{siunitx}
\usepackage{array}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{xspace}
\usepackage{relsize}
\usetikzlibrary{fadings}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Macros
\newcommand{\tocite}{\textbf{CITE}}
\newcommand{\quickcheck}{\emph{QuickCheck}\xspace}
\newcommand{\megadeth}{\emph{MegaDeTH}\xspace}
\newcommand{\dragen}{\textbf{DRAGEN}\xspace}

\newcommand{\evalrep}[2]{\ensuremath{\llbracket #1 \rrbracket_{#2}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some parameters
\setlength{\mathindent}{1.5\parindent}

@@

In the previous definition, we declare \ensuremath{\Conid{Exp}} as the sum of three possible
classes of values: \ensuremath{\Conid{Val}} represents literal values, whereas \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}
represent the addition and multiplication of two integer expressions,
respectively.
%
In Haskell, \ensuremath{\Conid{Val}}, \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}} are called \emph{data constructors} (or
constructors for short) and are used to distiguish which variant of the data
type we are reffering to.
%
Each data constructor is then defined as a product of zero or more types known
as \emph{fields}.
%
In particular, \ensuremath{\Conid{Val}} contains a field of type \ensuremath{\Conid{Int}}, while \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}
contain two fields of type \ensuremath{\Conid{Exp}} representing the operands of each operation.
%
Note that \ensuremath{\Conid{Exp}} is used as a field of at least one of its data constructors
(case \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}), making it a recursively defined ADT.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this reprentation, the expression ``2 + (5 * 6)'' can be encoded simply by
using \ensuremath{\Conid{Exp}} data constructors as \ensuremath{(\Conid{Add}\;(\Conid{Val}\;\mathrm{2})\;(\Conid{Mul}\;(\Conid{Val}\;\mathrm{5})\;(\Conid{Val}\;\mathrm{6})))}.
%
Furthermore, an evaluation function from \ensuremath{\Conid{Exp}}s to integer values can be defined
very idiomatically:

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}c<{\hspost}@{}}%
\column{6E}{@{}l@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}gen_{Exp}\mathrel{=}\Varid{sized}\;(\lambda \Varid{size}\to {}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{if}\;\Varid{size}\equiv \mathrm{0}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{then}\;\Conid{Val}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{else}\;\Varid{frequency}{}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{}[\mskip1.5mu {}\<[6E]%
\>[9]{}(\mathrm{1},\Conid{Val}{}\<[18]%
\>[18]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{2},\Conid{Add}{}\<[18]%
\>[18]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{Exp}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{Exp}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{1},\Conid{Mul}{}\<[18]%
\>[18]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{Exp}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{Exp})\mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

This random generator is defined using \quickcheck's combinator \ensuremath{\Varid{sized}} to
paremetrize the generation process up to an external integer number known as the
\emph{generation size}.
%
This parameter is used to limit the maximum amount of recursive call that this
random generator can perform.
%
When called with a positive generation size (case \ensuremath{\Varid{gen}\;\Varid{n}}), the generator will
pick among any data constructor of \ensuremath{\Conid{Exp}} with a explicitly given frequency---two
times more \ensuremath{\Conid{Add}}s than \ensuremath{\Conid{Val}} or \ensuremath{\Conid{Mul}}.
%
In the case it picks to generate a \ensuremath{\Conid{Val}} data constructor, it also generates a
random \ensuremath{\Conid{Int}} value using the standard overloaded generator \ensuremath{\Varid{arbitrary}}
(\quickcheck provides standard random generators for most base types like \ensuremath{\Conid{Int}},
\ensuremath{\Conid{Bool}}, etc.).
%
On the other hand, when it randomly picks to generate either an \ensuremath{\Conid{Add}} or a \ensuremath{\Conid{Mul}}
data constructor, it also generates independently two random subexpressions
corresponding to the data constructor fields by calling itself recursively
(\ensuremath{\Varid{resize}\;(n{-}1)\;gen_{Exp}}), decreasing the generation size on each recursive
invocation.

@@

\quickcheck's default implementation for random \ensuremath{\Conid{Int}}s pick a number in the
interval $[-n, n]$ with uniform distribution.
%
Hence, in principle we need to recognize what is a suitable generation size,
which should be big enough for our generator be able to generate the numbers we
pattern match against to, otherwise we will not be able to generate any value to
match against \ensuremath{\Varid{foo}} clauses, leaving fragments of code completely untested.
%
Suppose then that we pick the a generation size $50$, i.e. the minimum size that
is big enough to produce an \ensuremath{\Conid{Int}} number equal to $50$.
%
Under this consideration, the probability of generating a value matching the
first clause of \ensuremath{\Varid{foo}} (and hence triggering the first error) results as follows:
%
\begin{align*}
  &P(match(foo\#1))\\
  &\quad = P(Add)       * P(Add) * 1 * (P(Val) * (1/100)) \\
  &\phantom{xxxxxxxxxx} * P(Add) * (P(Val) * (1/100)) * 1
\end{align*}

If we use \megadeth to automatically derive a random generator for \ensuremath{\Conid{Exp}}, we
obtain a uniform generation probability distribution over constructors, i.e.,
$P(\ensuremath{\Conid{Val}}) = P(\ensuremath{\Conid{Add}}) = P(\ensuremath{\Conid{Mul}}) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of \ensuremath{\Varid{foo}} only once.
%
This situation can be somewhat improved if we use \dragen to obtain a random
generator.
%
Using this tool, we can optimize the generation probabilities in order to
benefit the generation of some data constructors over the rest.
%
Considering that the first pattern matching of \ensuremath{\Varid{foo}} involves the data
constructor \ensuremath{\Conid{Add}} as the only recursive one, we can set an target generation
proportion of \ensuremath{\Conid{Add}} data constructors of, for instance, $20:1$ with respect to
the rest of the generated data constructors.
%
By doing so, the obtained distribution of values results such that
$P(match(foo\#1)) \approx 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Html}\mathrel{=}\Conid{Html}\;\Conid{String}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{head}\mathbin{::}\Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{head}\;(\Conid{Html}\;\Varid{inner}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Conid{Html}\;(\text{\ttfamily \char34 <head>\char34}\plus \Varid{inner}\plus \text{\ttfamily \char34 </head>\char34}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{body}\mathbin{::}\Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{body}\;(\Conid{Html}\;\Varid{inner}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Conid{Html}\;(\text{\ttfamily \char34 <body>\char34}\plus \Varid{t}\plus \text{\ttfamily \char34 </body>\char34}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{br}\mathbin{::}\Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{br}\mathrel{=}\Conid{Html}\;\text{\ttfamily \char34 <br~/>\char34}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}(\mathop{\langle\! +\! \rangle})\mathbin{::}\Conid{Html}\to \Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}(\Conid{Html}\;\Varid{x})\mathop{\langle\! +\! \rangle}(\Conid{Html}\;\Varid{y})\mathrel{=}\Conid{Html}\;(\Varid{x}\plus \Varid{y}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

We believe that, if certain patterns of values are relevant enough to appear in
the codebase being tested, a practical random testing methodology should be able
to produce values satisfying such patterns in a substantial proportion.
%
The next section introduces a reprentation model that let us encode the
structure information presented here into our automatically derived random
generators in a modular and flexible way.
\section{Extracting Structure} \label{sec:hrep}

In this section we present a compositional representation to express the random
generation of values following the internal structure of their data types along
with the structure present on patterns matchings and abstract interfaces.

The key idea of this work is to represent different sources of structured value
information of a data type in a homogeneous way that we call a ``higher-level
representation'' (\ensuremath{\text{HRep}} from now on).%
\footnote{The notion of higher-level comes from that the generation process of
  given target data type is entirely determined by the type of the chosen
  representation, instead of by a concrete generator defined at the term level.}
%
For this purpose, we use a series of automatically derived data types, each one
representing an atomic unit of information that can be randomly generated and
then reflected back to the corresponding value of the original data type.

We wil reuse the previously defined data type \ensuremath{\Conid{Exp}} and the function \ensuremath{\Varid{foo}} to
explain the different concepts involved all across this section.

\subsection{\textbf{Representing Data Constructors}}

We begin by introducing the simplest representation that we can extract from our
codebase: the representation of single data constructor.
%
Each original constructor can be represented by a automatically derived data
type consisting of a single constructor with the same fields, except for the
recursive ones that are abstracted away.
%
In this light, we represent each constructor of the data type \ensuremath{\Conid{Exp}} as follows:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\text{HRep}_{Val}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}\text{Mk}_{Val}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{Add}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}\text{Mk}_{Add}\;\Varid{r}\;\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{Mul}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}\text{Mk}_{Mul}\;\Varid{r}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Note that the previous definition are type parametric over the type parameter
\ensuremath{\Varid{r}}.
%
This way we can replace \ensuremath{\Varid{r}} with any concrete data type, obtaining different
possible values on each case.
%
For instance, the value (\ensuremath{\text{Mk}_{Add}\;\mathrm{10}\;\mathrm{20}}) has type \ensuremath{\text{HRep}_{Add}\;\Conid{Int}}, while the
value (\ensuremath{\text{Mk}_{Mul}\;\Conid{True}\;\Conid{False}}) has type \ensuremath{\text{HRep}_{Mul}\;\Conid{Bool}}.
%
In practice, this parametricity let us instatiate \ensuremath{\Varid{a}} with the type of the
chosen representation (which might change according to the user demands),
without having to modify anything in the underlying machinery.

For this purpose, we employ Haskell's type classes mechanism \tocite{}.
%
We define an evaluation type class $(\Downarrow)$ which specifies how to perform
a single transformation \ensuremath{\Varid{step}} from an \ensuremath{\text{HRep}} back into a concrete value of the
target:
%
\begin{align*}
  \llbracket \_ \rrbracket_{target}\ :\ rep_{target}\ \rightarrow\ target
\end{align*}

% \begin{code}
% class (rep down target) where
%   step :: rep target -> target
% \end{code}

Then, we can define the following overloaded instances of the \ensuremath{\Varid{step}} operation
for the canonical \ensuremath{\text{HRep}} of data constructors simply by mapping each lifted
constructor back into its corresponding one.
%
\begin{alignat*}{4}
  &\evalrep{\ensuremath{\text{Mk}_{Val}}\ n\ \ \ &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Val}}\ n \\
  &\evalrep{\ensuremath{\text{Mk}_{Add}}\ x\ y   &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Add}}\ \evalrep{x}{\ensuremath{\Conid{Exp}}}\ \evalrep{y}{\ensuremath{\Conid{Exp}}} \\
  &\evalrep{\ensuremath{\text{Mk}_{Mul}}\ x\ y   &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Mul}}\ \evalrep{x}{\ensuremath{\Conid{Exp}}}\ \evalrep{y}{\ensuremath{\Conid{Exp}}}
\end{alignat*}

% \begin{code}
% instance (HRep_Val down Exp) where
%   step (Mk_Val n) = Val n

% instance (HRep_Add down Exp) where
%   step (Mk_Add x y) = Add x y

% instance (HRep_Mul down Exp) where
%   step (Mk_Mul x y) = Mul x y
% \end{code}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{Val}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}\text{Mk}_{Val}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[E]%
\\
\>[3]{}gen_{Add}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}\text{Mk}_{Add}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\\
\>[3]{}gen_{Mul}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}\text{Mk}_{Mul}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

\subsection{\textbf{Type Level Combinators}}

In order to compose the \ensuremath{\text{HRep}}s for the single data constructors already
presented, we will need to define a series of type level combinators.

In first place, we define a type combinator ($\otimes$) to tag a \ensuremath{\text{HRep}} with an
explicit generation frequency:

This combinator is evaluated back to our target data type simply by piping the
result from the inner representation.
%
It does not change the evaluation semantics, as it is only considered at
generation time:
%
\begin{align*}
  \evalrep{Freq\ x : f \otimes n}{t} = \evalrep{x : f}{t}
\end{align*}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}gen_{f \otimes n}\;gen_{r}\mathrel{=}\Conid{Term}\mathop{\mathsmaller{\langle \$ \rangle}}gen_{f}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

% \begin{code}
% instance (f otimes n down Exp) where
%   step (Freq f) = step f
% \end{code}

% \begin{code}
% instance Arbitrary1 (f otimes n) where
%   liftArbitrary gen_HRep = Tag <$> gen_HRep
% \end{code} %$

When generating random values, we will only allow certain terminal
representations to be generated at the moment that the generation size gets
exhausted.
%
For this, we follow the same approach as before to define a combinator (\ensuremath{\_^\bullet})
to tag terminal representations.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;(f^\bullet)\;\Varid{a}\mathrel{=}\Conid{Term}\;(\Varid{f}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{align*}
  \evalrep{Term\ x : \ensuremath{f^\bullet}}{t} = \evalrep{x : f}{t}
\end{align*}
